---
title: "cloud native access"
output: github_document
date: '2022-05-05'
---

```{r setup}
library(tidyverse)
library(arrow)
```

All big commercial cloud providers (e.g. AWS, Google, Azure) provide object stores. Amazon's AWS S3 platform is among the most widely used, and hosts a [large registry of open datasets](https://registry.opendata.aws/), including archives of NOAA global circulation models.  A wide range of software tools have been built around the AWS S3 interface, which has led to the adoption of S3-compatible interfaces by other cloud providers (Google) and open source platforms (CEPH, MINIO). While not an official community standard like POSIX, this makes S3 something of a de-facto standard.

EFI uses the Jetstream2 Object Store, maintained by NSF XSEDE program. Unlike commercial providers, Jetstream2 uses an open source system called [CEPH](https://docs.ceph.com/en/quincy/) (developed by RedHat), which supports the S3 interface. 

Here we'll use the popular [Apache Arrow](https://arrow.apache.org/) framework 


```{r}
s3 <- s3_bucket("forecasts", endpoint_override = "https://")
```