---
title: "cloud native access part 2: spatially explicit data"
output: github_document
date: '2022-05-010'
---


# Synopsis

In part I, we looked at how we could access parts of tabular data files using a cloud-native protocol.
See README.md for an overview of the motivation behind these approaches.



Here's a dummy example using Sentinel geotifs and computing NDVI. 
We could replace with a more "EFI" example using NOAA GEFS forecasts instead.  

```{r setup, message=FALSE}
library(tidyverse)
library(rstac)
library(terra)
library(sf)
```


```{r}
## could replace with a VSIZIP / VSICURL example 
san_fran <- st_read("https://dsl.richmond.edu/panorama/redlining/static/downloads/geojson/CASanFrancisco1937.geojson")
bbox = st_bbox(san_fran)
```


```{r}
## Search for data from a given collection in a given bounding box:
s_obj <- stac("https://planetarycomputer.microsoft.com/api/stac/v1/")

it_obj <- s_obj |> 
  stac_search(collections = "sentinel-2-l2a",
              bbox = bbox,
              datetime = "2019-06-01/2019-08-01"
              ) |>
  get_request() |> # intersects needs post
  items_sign(sign_fn = sign_planetary_computer())

properties <- map(it_obj$features, "properties")

## filter using metadata to find image in search match with fewest clouds
cloud_cover <- map_dbl(properties, "eo:cloud_cover")
best <- which.min(cloud_cover)
best_img <- it_obj$features[[best]]

download_url <- best_img$assets$visual$href

```

Now time for some virtual filesystem magic!  Instead of downloading the raster, we put the magic prefix `/vsicurl/` _in front_ of the URL, and pass that directly to `rast()`!

```{r}
data <- terra::rast(paste0("/vsicurl/", download_url)) ## lazy-read ~ 100 MB file over the network, no disk storage
plot(data)
```


We can do subsets and computation over the network too. Here we calculate NDVI using a subset of bands.  

```{r}
## ugh, bbox must be re-projected into correct units for Sentinel
bb <- bbox %>% st_as_sfc() %>%
  st_transform(crs = st_crs(crs(data))) %>%
  st_bbox() 
terra_bbox <- ext(bb$xmin, bb$xmax, bb$ymin, bb$ymax)


## access red and infra-red bands to compute NDVI
red <- terra::rast(paste0("/vsicurl/", best_img$assets$B04$href)) |> crop(terra_bbox) ## lazy-read ~ 100 MB file over the network, no disk storage
nir <- terra::rast(paste0("/vsicurl/", best_img$assets$B08$href)) |> crop(terra_bbox) ## lazy-read ~ 100 MB file over the network, no disk storage

## see https://rspatial.org/terra/rs/3-basicmath.html
vi2 <- function(x, y) {
  (x - y) / (x + y)
}
ndvi <- lapp(c(nir, red), fun = vi2)

# and here we are!
ndvi  |> plot()
```
